Developing an automated iterative near-term forecasting system for an ecological

### 2018-07-02 manuscript/response_to_review.Rmd

## Associate Editor

**The main issue regards the motivation (reviewer 1) and target audience (reviewer 2) of the application. If this is meant for researchers that are proficient at modeling and coding, it should be made clearer how they benefit from this automated application (wouldn't they already have similar protocols and pipelines in place in their routine research?). If instead, this is for empiricists, or environmental managers, with little coding experience, the manuscript is too technical and the application perhaps should be made more 'user-friendly' (reviewer 1). My suspicion (and reviewer 2's) is that it aims for that increasingly common middle person (the computationally literate ecologist). However, as Reviewer 2 points out, it is not clear how such audience should interact with the defined pipeline and how, for example, they could propose new models. A fourth (very attractive) possibility is that this is meant as an interface for people with different skills and motivations to collaborate (e.g. to foster the feedback between scientific development and applied use). If this is the case, the protocols of interaction need to be clarified, as well as the audience for this article (e.g. is it the computational ecologist who will read this article and propose applied scientists to work with her/him through this system?). Defining the audience clearly should guide how to improve the application and manuscript in a way that it ensures its use and long-term survival as an important short-term forecasting tool.**

We have expanded the last paragraph of the Introduction to address the question of target audience and split the opening paragraph of the Introduction into two paragraphs to provide expanded treatment of the motivation for this approach including the addition of three examples.

The AE and reviewer are correct that we have target this paper primarily at the "increasingly common middle person" and "researchers that are proficient at modeling and coding". Thanks to the expanded efforts at computational training, we believe that many researchers have the necessary skills to implement these sorts of approaches, and we think that the association between "empiricists" and "little coding experience" is increasingly out of date as demonstrated by the fact that this system was developed in large part by the empiricists collecting the data.

That said, we certainly recognize that not all ecologists and evolutionary biologists have the necessary skills to implement this sort of pipeline (just as many do not have the skills to implement a Bayesian statistical analysis). As a result, a single paper in this space won't be perfect for everyone. We have tried hard to reach a middle ground where users with the necessary skills to implement this sort of system will be able to understand precisely what we did while also providing introduction and motivation for researchers with less advanced computational skills to learn more and collaborate. To reach researchers with relatively basic computational skills, we have recently completed a paper focused on the data management component of this pipeline that focuses on providing a ground up introduction to this kind of automation in an ecological context (https://doi.org/10.1101/344804). We hope that the combination of these two papers will successfully introduce this general class of approaches and concepts to a broad array of researchers.

We disagree with the assumption that researchers that are proficient at modeling and coding would "already have similar protocols and pipelines in place in their routine research." We are active in this field and aware of very few efforts of this scope in ecology. These researchers will have the foundational tools to develop similar systems (or contribute to this one).

@@

There is definitely a need and demand for short-term forecasting. This was laid out in detail in Dietze et al. (2018; PNAS), but we agree that the current paper benefits from providing more direct treatment of this. We have significantly expanded the opening paragraphs of the introduction to provide specific examples and further emphasize the value of this approach.

**I might also suggest differentiating parameter estimation and model development. Forecasts can improve by adding new models, or by adding new data which then improves parameter estimation in the old models.  The automatic pipeline is great at the second part of this, continually adding data and re-fitting the old models to the new data and then producing new forecasts.  However, since there was so much data in the Portal project to begin with, I doubt that parameter estimates in the models have changed much since the automatic pipeline was established (maybe they have?).  However, in a new project that was just getting started (e.g. NEON), the main improvements in forecast skill early on would be in adding more data.  Eventually, however, the models might stop improving and reach some prediction limit imposed by the structure of the model.  As the authors point out in the discussion, finding this point will be key for improving ecological theory. I think a lot of ecologists think their models are useful, they just need better parameter estimates.  My sense is that actually, most ecological models are just crap for forecasting, it’s not for lack of data! Maybe this is an obvious point, but the role of improved parameter estimation in this whole endeavor kind of gets lost for me.  Does the pipeline even save the fitted model parameters? Would that be worthwhile?**

There are a number of interesting points in this paragraph that we feel are beyond the scope of this current paper. We have already described the importance of new models and how we have facilitated adding them. We have now further expanded this section to describe how those external to the group can contribute improved models. Introducing and discussing the different implications of different forms of error and how the are associated with increasing in data vs. improvements in models is an interesting subject for forecasting but is out of scope of this paper. It would definitely be something interesting to explore in future work.

@@

**In Figure 1, I feel that something is missing here about how other scientists are supposed to interact with the data and forecasts.  The manuscript makes it pretty clear that this is not just for internal use, but is it actually inviting other labs to add models and make predictions? How specifically would another lab contribute a model to the project? Would they fork the project on github, develop a new model and then do a pull request? Or would they create a separate prediction archive themselves? Maybe add an arrow in Fig 1. showing how new models can be plugged into the forecasting work flow.  Box 1 addresses some of this but I guess I want more details.**

We have added description of how others can contribute to the Modeling and Forecasting section of the paper. We tried to also add this to the figure, but could not find a way to do so without making the figure overly crowded and complicated.

**Some other miscellaneous questions:**
**Can models be dropped from the ensemble?**

This is not currently done at the moment but is on the development roadmap.

@@

**Lines 63 to 83:  Could this section include some examples that demonstrate the value of iterative near-term forecasting from other fields? Could it be argued that iterative near term forecasting has led to progress in other fields? Say meteorology and political science?**

We have now mentioned the established benefits for iterative near term forecasting in meteorology and added several citations for readers interested in exploring this further.

**Figure 1: Another point on figure one.  What about model improvement? It seems like there should be a feedback loop somewhere showing how models get refined, added, dropped etc.**

We tried to add a visual representation of this to Figure 1 but could not make it work without making the figure overly crowded. We have added some additional discussion of this "Modeling and Forecasting" section of the paper.

@@

**- A, B are not labeled in the figure**

Fixed.

**This paper is written to quite a narrow audience: ecologists that already code and could benefit from this kind of automation. This is fine. But many researchers do not yet have the coding skills to benefit from automation (the audiences for Software Carpentry, who the authors mention). To me, it seems that in ecology there are two computational camps at the moment: ecologists who are ready for the workflows mentioned in this paper, and those beginning to engage through Software Carpentry. This is obviously over-simplified, but as I read this manuscript I found myself thinking that it would be nice to try to speak to both camps, instead of just the one with the skills for this automation. How do people get there? How do we get ecologists to the point where they too can develop a workflow like the one in this manuscript?**

A single paper in this space won't be perfect for everyone. We have tried hard to reach a middle ground where users with the necessary skills to implement this sort of system will be able to understand precisely what we did while also providing introduction and motivation for researchers with less advanced computational skills to learn more and collaborate. Thankfully, given recently expanded efforts to provide more advanced computational training to scientists, we disagree that it is currently written for "quite a narrow audience". That said, we agree with the reviewer that reaching out to researchers with relatively basic computational skills is important. To address this broader need for introductory material on automation and continuous integration in ecology, we have recently completed a paper focused on the management of data that is regularly updated, which focuses on communicating at an introductory level, including the inclusion of a tutorial for helping beginners to this area get started (https://doi.org/10.1101/344804). We hope that the combination of these two papers will successfully introduce these approaches and concepts to a broad array of researchers.

@@

**I was happy to see the discussion of team science (~L421), but training and skill-building immediately came to mind, and I didn't know who the "team" was, it seems that it is the Weecology team, not including the Portal Project. Can the Portal Project maintain this on their own? I still didn't have a sense whether Florida researchers created something for Arizona field scientists that will be maintained by the Florida scientists, or whether the Arizona group was trained to maintain it on their own. My assumption is that the Arizona team would be in the Software Carpentry camp, and I'm thinking about how to bridge these camps and empower more scientists with the skills to be able to understand the integrative process the manuscript describes. The goal is to get all ecologists into the automation camp, how do we do it?**

**I would suggest adding some sentences that address this to the intro and discussion. This could traditionally be seen as out of the scope of the manuscript, but that will continue to perpetuate the two camps instead of bringing the Software Carpentry camp towards the automation camp.**

As is already described in the paper, the field researchers are an active part of the team that develops and runs this system. Both components of the team are part of Weecology. As we've mentioned above, the clear distinction between empiricists and computional folks really isn't valid anymore and the field researchers involved in this project all have strong computational skills. We have modified the language in this section to clarify how the team works. While we agree that "The goal is to get all ecologists into the automation camp", and we are working hard on multiple fronts to make this a reality, this is a larger challenge that is beyond the scope of this current paper.

### 2018-07-02 manuscript/portal_iterative_forecasting_ms.Rmd
Update portal_iterative_forecasting_ms.Rmd

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of how well models capture the processes governing ecological systems [@clark2001; @tallis2006; @diaz2015; @dietze2017]. In 2001, Clark et al. [-@clark2001] called for a more central role of forecasting in ecology. Since then, an increasing number of ecological forecasts are being published that focus on a societally important questions from daily to decadal time scales [@dietze2018]. At daily scales, ecological forecasts predict the occurance of environmental issues like toxic algal blooms [@stumpf2009]. At quarterly scales forecasts are used to predict the stocks of fisheries and make decisions about quotas for fishing [@NOAA2016]. At decadal time scales, ecological forecasts are used to predict how biodiversity will change as it responds to anthropogenic influences [@harris2018]. These forecasting examples highlight the important role that ecological forecasts play in recasting ecological knowledge in societally relevant ways and also improve our understanding of ecological systems by testing the ability of our models to predict how systems will change in the future [@dietze2018, @harris2018].

@@

With the goal of making ecological forecasting more dynamic and responsive, Dietze et al [-@dietze2018] recently called for an increase in iterative near-term forecasting. Iterative near-term forecasting is defined as making predictions for the near future and repeatedly updating those predictions through a cycle of evaluation, integration of new data, and generation of new forecasts. Because forecasts are made 'near-term'---daily to annual time scales instead of multi-decadal---predictions can be assessed more quickly and frequently, leading to more rapid model improvements [@tredennick2016; @dietze2018]. Since forecasts are made repeatedly through time, new data can be continuously integrated with each iteration [@dietze2018]. By quickly identifying how models are failing, facilitating rapid testing of improved models, and incorporating the most up-to-date data available, iterative near-term forecasting has the potential to promote rapid improvement in the state of ecological forecasting. In addition to yielding improved information for guiding policy and management [@clark2001; @luo2011; @petchey2015], this iterative approach will help improve our basic understanding of ecological systems [@dietze2018]. For example, alternative mechanistic models can be compared to determine which model provides the best forecasts, thus providing insights into the importance of different ecological processes [@dietze2018]. Iterative near-term forecasting provides the more dynamic interplay between models, predictions, and data that has been identified as necessary for improving ecological forecasting and our understanding of ecological systems more broadly.

Because iterative near-term forecasting requires a dynamic integration of models, predictions, and data, Dietze et al [-@dietze2018] highlight approaches to data management, model construction and evaluation, and cyberinfrastructure that are necessary to effectively implement this type of forecasting (Box 1). Data needs to be released quickly under open licenses [@vargas2017; @dietze2018] and structured so that it can be used easily by a variety of researchers and in multiple modeling approaches [@borer2009; @strasser2011]. Models need to be able to deal with uncertainty, in both the predictors and the predictions, to properly convey uncertainty in the resulting forecasts [@diniz-filho2009]. Multiple models should be developed, both to assess which models are performing best [@dietze2018] and to facilitate combining models to form ensemble predictions which tend to perform better than single models [@araujo2007; @diniz-filho2009]. Ensuring that data and models are regularly updated and new forecasts are made requires cyberinfrastructure to automate data processing, model fitting, prediction, model evaluation, forecast visualization, and archiving. In combination, these approaches should allow forecasts to be easily rerun and evaluated as new data becomes available [Box 1; @dietze2018].

While iterative near-term forecasting is an important next step in the evolution of ecological forecasting, the requirements outlined by Dietze et al (Box 1) are not trivial to implement (e.g., making quality data available in near real-time and automatically rerunning forecasts in reproducible ways), and few of their recommendations are in widespread use in ecology today [e.g., @wilson2014; @stodden2014; @yenni2018]. We explored what it would entail to operationalize Dietze et al's recommendations by constructing our own iterative near-term forecasting pipeline for an on-going, long-term ecological study that collects high-frequency data on desert rodent abundances [@brown1998; @ernest2008]. We constructed an automated forecasting pipeline with the goal of being able to forecast rodent abundances and evaluate our predictions on a monthly basis. In this paper, we discuss our approach for creating this iterative near-term forecasting pipeline, the challenges we encountered, the tools we used, and the lessons we learned so that others can create their own iterative forecasting systems. For those interested in implementing iterative forecasting, either on their own or as part of a team, this paper will provide a roadmap for how to build such a system and what skills will be helpful to do so. For readers looking for an introduction to automation and continous integration in an ecological context, we recommend our paper on data management for continuously collected data, which includes a tutorial on how to set up some of the aspects of automation described in this paper (@yenni2018).

@@

Iterative forecasting is most effective with frequently collected data, since it provides more opportunities for updating model results and assessing (and potentially improving) model performance [Box 1; @dietze2018]. The Portal Project is a long-term ecological study situated in the Chihuahuan Desert (2 km north and 6.5 km east of Portal, Arizona, US). Researchers have been continuously collecting data at the site since 1977, including data on the abundance of rodent and plant species (monthly and twice yearly, respectively) and climatic factors such as air temperature and precipitation (daily) [@brown1998; @ernest2009; @ernest2016]. The site consists of 24 50m x 50m experimental plots. Each plot contains 49 permanently marked trapping stations laid out in a 7 x 7 grid, and all plots are trapped with Sherman live traps for one night each month. For all rodents caught during a trapping session, information on species identity, size, and reproductive condition is collected, and new individuals are given identification tags. This information on rodent populations is high-frequency, uses consistent trapping methodology, and has an extended time-series (470 monthly samples and counting), making this study an ideal case for near-term iterative forecasting.

@@

Forecasts are currently evaluated using root mean square error (RMSE) to evaluate point forecasts and coverage to evaluate uncertainty. We plan to add additional metrics in the future. In addition to evaluating the actual forecasts, we also use hindcasting [forecasting on already collected data; @jolliffe2003] to gain additional insight into the methods that work best for forecasting this system. For example, a model is fit using rodent observations up to June 2005, then used to make a forecast 12 months out to May 2006. The observations of that 12-month period can immediately be used to evaluate the model. Since hindcasting is conducted using data that has already been collected, it allows model comparisons to be conducted on large numbers of hindcasts and provides insight into which models make the best forecasts without needing to wait for new data to be collected [@harris2018]. It can also be used to quickly evaluate new models instead of waiting for an adequate amount of data to accumulate. As the performance of different models is understood through evaluation of forecasts and hindcasts models can be refined or removed from the system or ensemble to iteratively improve the resulting forecasts.

>>>>

Forecasting the future state of ecological systems is important for management, conservation, and evaluation of how well models capture the processes governing ecological systems [@clark2001; @tallis2006; @diaz2015; @dietze2017]. In 2001, Clark et al. [-@clark2001] called for a more central role of forecasting in ecology. Since then, an increasing number of ecological forecasts are being published that focus on a societally important questions from daily to decadal time scales [@dietze2018]. At daily scales, ecological forecasts predict the occurance of environmental issues like toxic algal blooms [@stumpf2009]. At quarterly scales, forecasts are used to predict the stocks of fisheries and make decisions about quotas for fishing [@NOAA2016]. At decadal time scales, ecological forecasts are used to predict how biodiversity will change as it responds to anthropogenic influences [@harris2018]. These forecasting examples highlight the important role that ecological forecasts play in recasting ecological knowledge in societally relevant ways and also improve our understanding of ecological systems by testing the ability of our models to predict how systems will change in the future [@dietze2018, @harris2018].

@@

With the goal of making ecological forecasting more dynamic and responsive, Dietze et al. [-@dietze2018] recently called for an increase in iterative near-term forecasting. Iterative near-term forecasting is defined as making predictions for the near future and repeatedly updating those predictions through a cycle of evaluation, integration of new data, and generation of new forecasts. Because forecasts are made 'near-term'---daily to annual time scales instead of multi-decadal---predictions can be assessed more quickly and frequently, leading to more rapid model improvements [@tredennick2016; @dietze2018]. Since forecasts are made repeatedly through time, new data can be continuously integrated with each iteration [@dietze2018]. By quickly identifying how models are failing, facilitating rapid testing of improved models, and incorporating the most up-to-date data available, iterative near-term forecasting has the potential to promote rapid improvement in the state of ecological forecasting. In addition to yielding improved information for guiding policy and management [@clark2001; @luo2011; @petchey2015], this iterative approach will help improve our basic understanding of ecological systems [@dietze2018]. For example, alternative mechanistic models can be compared to determine which model provides the best forecasts, thus providing insights into the importance of different ecological processes [@dietze2018]. Iterative near-term forecasting provides the more dynamic interplay between models, predictions, and data that has been identified as necessary for improving ecological forecasting and our understanding of ecological systems more broadly.

Because iterative near-term forecasting requires a dynamic integration of models, predictions, and data, Dietze et al. [-@dietze2018] highlight approaches to data management, model construction and evaluation, and cyberinfrastructure that are necessary to effectively implement this type of forecasting (Box 1). Data needs to be released quickly under open licenses [@vargas2017; @dietze2018] and structured so that it can be used easily by a variety of researchers and in multiple modeling approaches [@borer2009; @strasser2011]. Models need to be able to deal with uncertainty, in both the predictors and the predictions, to properly convey uncertainty in the resulting forecasts [@diniz-filho2009]. Multiple models should be developed, both to assess which models are performing best [@dietze2018] and to facilitate combining models to form ensemble predictions which tend to perform better than single models [@araujo2007; @diniz-filho2009]. Ensuring that data and models are regularly updated and new forecasts are made requires cyberinfrastructure to automate data processing, model fitting, prediction, model evaluation, forecast visualization, and archiving. In combination, these approaches should allow forecasts to be easily rerun and evaluated as new data becomes available [Box 1; @dietze2018].

While iterative near-term forecasting is an important next step in the evolution of ecological forecasting, the requirements outlined by Dietze et al. (Box 1) are not trivial to implement (e.g., making quality data available in near real-time and automatically rerunning forecasts in reproducible ways), and few of their recommendations are in widespread use in ecology today [e.g., @wilson2014; @stodden2014; @yenni2018]. We explored what it would entail to operationalize Dietze et al's recommendations by constructing our own iterative near-term forecasting pipeline for an on-going, long-term ecological study that collects high-frequency data on desert rodent abundances [@brown1998; @ernest2008]. We constructed an automated forecasting pipeline with the goal of being able to forecast rodent abundances and evaluate our predictions on a monthly basis. In this paper, we discuss our approach for creating this iterative near-term forecasting pipeline, the challenges we encountered, the tools we used, and the lessons we learned so that others can create their own iterative forecasting systems. For those interested in implementing iterative forecasting, either on their own or as part of a team, this paper will provide a roadmap for how to build such a system and what skills will be helpful to do so. For readers looking for an introduction to automation and continous integration in an ecological context, we recommend our paper on data management for continuously collected data, which includes a tutorial on how to set up some of the aspects of automation described in this paper (@yenni2018).

@@

Iterative forecasting is most effective with frequently collected data, since it provides more opportunities for updating model results and assessing (and potentially improving) model performance [Box 1; @dietze2018]. The Portal Project is a long-term ecological study situated in the Chihuahuan Desert (2 km north and 6.5 km east of Portal, Arizona, US). Researchers have been continuously collecting data at the site since 1977, including data on the abundance of rodent and plant species (monthly and twice yearly, respectively) and climatic factors such as air temperature and precipitation (daily) [@brown1998; @ernest2009; @ernest2016]. The site consists of 24 50m x 50m experimental plots. Each plot contains 49 permanently marked trapping stations laid out in a 7 x 7 grid, and all plots are trapped with Sherman live traps for one night each month. For all rodents caught during a trapping session, information on species identity, size, and reproductive condition is collected, and new individuals are given identification tags. This information on rodent populations is high-frequency, uses consistent trapping methodology, and has an extended time-series (475 monthly samples and counting), making this study an ideal case for near-term iterative forecasting.

@@

Forecasts are currently evaluated using root mean square error (RMSE) to evaluate point forecasts and coverage to evaluate uncertainty. We plan to add additional metrics in the future. In addition to evaluating the actual forecasts, we also use hindcasting [forecasting on already collected data; @jolliffe2003] to gain additional insight into the methods that work best for forecasting this system. For example, a model is fit using rodent observations up to June 2005, then used to make a forecast 12 months out to May 2006. The observations of that 12-month period can immediately be used to evaluate the model. Since hindcasting is conducted using data that has already been collected, it allows model comparisons to be conducted on large numbers of hindcasts and provides insight into which models make the best forecasts without needing to wait for new data to be collected [@harris2018]. It can also be used to quickly evaluate new models instead of waiting for an adequate amount of data to accumulate. As the performance of different models is understood through evaluation of forecasts and hindcasts, models can be refined or removed from the system or ensemble to iteratively improve the resulting forecasts.
